Applying State-free Reversible VAMPNets and Markov State Models to Learn Dynamics of DNA Oligonucleotides

### 2020-10-01 manuscript.tex
Updates from Overleaf

%Throughout this manuscript, we adopt "slithering" from previous work to describe dynamics between out-of-register states. We describe the metastable states along these pathway as "shifted."

In our analysis, we found that the AT-all sequence, given its repetitive structure and lack of GC-content, produced the cleanest dynamics and displayed a canonical spectral gap between modes. For this reason, we lead our discussion with this sequence and use it as a case study to work through our SRV-MSM pipeline step-by-step. Our first task was to identify the SRV lag time that was longer than the intrinsic Markov timescales of the system, yet short enough to resolve the dynamics of interest \citep{Phys2018MarkovValidation}. We found that most implied timescales converge at an SRV lag time of 1.2 ns. We kept a looser constraint on the convergence of the leading mode -- corresponding to the overall hybridization/dissociation process -- as it tends to have a longer transit period and becomes more difficult to fulfill Markovian conditions. It should be noted that our lag time selection was informed by the other sequences as well to maintain consistency. Next we selected an optimal number of SRV components to include in our analysis. After a certain point, higher order dynamical modes provide diminishing contributions the overall kinetic variance as measured by the Vamp-2 score, and the model can begin fitting on statistical noise in the trajectory data instead of the true dynamics \citep{McGibbon2015VariationalKinetics}. It is also more difficult to perform kmeans clustering on a high dimensional space, especially when those higher dimensions are less kinetically relevant \citep{Pande2010EverythingAsk}. For these reasons, the number of slow SRV components should be carefully selected based on the specific system of interest. As shown in (Figure \ref{fig:allseq_srv_crossval}), we see diminishing returns in the VAMP-2 score after five slow modes and select these modes as our optimized SRV basis. Next, we seek to interpret the physical relevance of these leading modes by plotting the Pearson correlation of each mode with the 100 intermolecular distances between strands. The quantitative meaning of these coordinates can be difficult to interpret given their nonlinear relationship to the SRV collective variables, but the relative different between these correlations shows which coordinates are most affected by (or effective on) each process. For example, the first slow mode shows a significant positive correlation to each distance and the strongest correlation with matching base pair distances (shown along the main diagonal). Given these relationships and the substantially longer timescale of this process, we can deduce that this leading mode corresponds to the dynamics of the overall hybridization and dissociation process. The next four SRV components all show a relatively high correlation along different offset diagonals. These diagonals correspond to the intermolecular distances between shifted base pairs and point to the existence of gradually faster "slithering" processes. Slow slithering mechanisms have similarly been reported in simulation studies to occur on orders of magnitude longer timescales than underlying fast dynamics such as fraying \citep{Markegard2015, Xiao2019, Maciejczyk2014DNAModel}.

@@

From this analysis we can determine that the slowest dynamics are fully characterized by the association/dissociation process and slithering behavior of out of register base pairs. Although this is qualitatively informative, we can access a more holistic picture of sequence kinetics and thermodynamics by using these SRV coordinates as a basis on which to construct an MSM. Because these coordinates are already capturing a majority of the system's kinetic variance, they serve as an ideal basis on which to group frames into microstates. We performed k-means clustering, and optimized the number of microstates at 200 by monitoring VAMP-2 score. Next, we selected an MSM lag time in a similar fashion to our SRV lag time selection process. In Figure \ref{fig:AT-all_dynamic} we compare the convergence of SRV-MSM and TICA-MSM timescales, where SRV-MSM timescales converge consistently faster and to higher values of all leading modes. This enables us to select a shorter lag time and build a higher resolution model than we could from an analogous TICA basis. Setting the MSM lag time to the same 1.2 ns we used for our SRVs, we built a Bayesian MSM to calculate transition probability matrix between each microstate. Finally, PCCA+ spectral clustering was implemented to group these microstates into macrostates that each represented a collection of metastable structures. Previous works have used a common set of microstates and/or performed manual clustering of microstates based on physical read outs from simulation data (stacking score, energies, etc) \citep{Pinamonti2017PredictingModels, PinamontiTheModels}. Although these techniques are useful for performing comparisons between sequences, we saw better results when optimizing MSMs to capture the most detail of sequence individually and thus developed an independent set of microstates and macrostates for each sequence. For AT-all, we kept to the convention of clustering into n+1 macrostates, where n is the number of slow components captured by the MSMs. To visualize these six macrostates, we project the data into the two leading TICA coordinates. Although SRVs outperform these coordinates for the purpose of MSM construction, TICA coordinates represent good high variance collective variables on which to visualize free energies and state assignments \citep{Sidky}. It is clear in Figure \ref{fig:AT-all_in_one} that the PCCA state assignments are capturing free energy minima in TICA space as independent states. After assigning these macrostates we can then calculate their relative probabilities and free energies, visualize representative molecular renderings, and estimates transition probabilities between states. In this final step, we use a minimum flux cutoff of 2e-6 in order to mitigate erroneous quick transition or skipping between states.

@@

Beyond state probabilities and free energy approximations, the macrostate MSM yields valuable discrete kinetic information in the form of transition probabilities between states. Figure \ref{fig:AT-all_in_one} shows the probability of moving from one state to another within the MSM lag time (1.2 ns). All transition probabilities are higher when moving towards a more aligned state than towards a more shifted state, suggesting that these metastable shifted states play a more significant role in facilitating the hybridization process than dissociation. Furthermore, we see equal or higher transition rates from shifted states to the dissociated state than to more aligned states, indicating that the slithering-hybridization process is frequently disrupted by complete dissociation. In particular, we observe that the transition probability from double-shifted 3' state to the dissociated state is 6x higher than to the neighboring single-shifted 3' state. Although there is an approximately equal probability of transitioning from the dissociated state to the hybridized and 5' shifted states, there is a 2x lower probability of moving to the 3' shifted states. These asymmetric kinetic effects, in addition to NN and inert tail contributions described above, may further lower the 3' shifted states populations relative to 5'. From these observations we conclude that shifted states are important in facilitating the hybridization process, but that increasingly shifted states (in particular when shifted in the 3' direction) are more likely to dissociate than to evolve into a fully aligned hybridized state. We recognize that these effects could be sensitive to the specific ion environment -- especially considering the effect of ionic strength on inert tails -- and warrant further computational and experimental investigation beyond this work.

@@

Next we examine the GC-end sequence, which add G:C caps to the same repetitive AT motifs. Based on SRV cross-validation (SI) we use the leading four slow modes to define our MSM basis.  These coordinates are qualitatively similar to those we studied for AT-all, where the three faster modes each represent some slithering dynamics. When we build the SRV-MSM for the GC-end sequence we observe a larger separation between the leading SRV-MSM timescale and the proceeding modes, which is shown in SI Figure \ref{fig:GC-end_dynamic}) by a distinct spectral gap. We observe a second spectral gap after the fourth slow modes along with a drop in the model vamp-2 score (SI Figure \ref{fig:srv_crossval}). The clustered macrostates show a similar distribution where shifted populations are notably lower. The 3' double-shifted state is no longer identified as a metastable state, and the total number of states is reduced to five (again one more than the total number of slow modes). Although these states appear similar to the AT-all shifted states, the presence of the C:T or G:A mismatches replace two intact WC bonds in the corresponding AT-all states. This substantially diminishes the stability of the GC-end shifted states, resulting in lower state populations and shorter timescales.

@@

The GC-core sequence represents a departure from the dominant slithering dynamics observed for AT-all and GC-end. Based on SRV cross-validation (SI Figure \ref{fig:srv_crossval}), we selected the first three components for our analysis. Next, we built an SRV-MSM using these first three SRV modes as a basis and proceeded along the pipeline as described above. We found that four macrostate clustering was unstable -- likely because the 3rd mode is mostly providing  information about dissociation dynamics -- so we performed PCCA+ clustering into three macrostates representing the hybridized, dissociated, and "4bp-frayed" states. Again, we found transition probabilities between states and visualized representative molecular renderings. We observe that the 4bp-frayed meta-stable state is solely composed of trajectories where both G:C core base pairs are bound and one of the adjacent A:T bonds are not. Previous work suggests that once key contacts are made, the zippering mechanism ensures that the helix will quickly form outward \citep{Romano2013DNADependence, Yin2011KineticsHybridization}. Our results indicate, however, that the relative instability of AT bonds compared the the GC-core can interrupt this process and form a longer lived frayed metastable states. This occurs during the dissociation process as well, where one half of the A:T base contacts are entirely broken for a substantial period of time before the full dissociation event occurs. We observe these events to occur with equal probability on either permutable side of the helix. The MSM transition probabilities suggest that this is key intermediate state for both the hybridization and dissociation processes, however the pathways differ between directions. We find that the hybridized state has a 5x higher probability of transitioning into the frayed state within the lag time compared to transitions from the dissociated state. This is expected considering that this state is more accessible from an already bound helix. Moreover, once oligos are in this state, they are over 10x more likely to return to the hybridized state than to fully dissociate. Thus once a dissociated to 4bp-frayed transition has occurred, it is a likely to proceed into a fully hybridized conformation. On the other hand, transitions from the hybridized to 4bp-frayed state are much more frequent and are unlikely to proceed to a fully dissociated state.

@@

For GC-core, each base pair and cutoff combination appears linear with temperature, indicating a barrierless and diffusion-driven process. As expected, the acceleration factor is also highly dependent on the choice of parameters, but, given that terminal fraying likely dominates the signal, we approximate that fraying dynamics are accelerated by about two order of magnitude relative to experiments. In contrast, GC-end is distinctly exponential, especially at base pairs closer to the core of the duplex. This behavior makes sense given the thermodynamic barrier to G:C fraying, but it also deviates the GC-end experimental linear fit. In addition to fraying, we notice that the rate of slithering increases at higher temperature. Moreover, the thermodynamic stability of these shifted conformations increases relative to that of intact duplex. For T-jumps at higher initial and final temperatures, it is possible that some portion of the starting ensemble is in a shifted state (cite Ryan's work) and that the fast dynamics are increasingly influenced by alternative "looping" G:C bonding that we observed in our diffusion map analysis above. It is less likely that shifting mechanisms would have their own distinct signals, given that these occur at far lower frequency than fraying or full-dissociation, but we believe that further experimental work is necessary to investigate this phenomenon.

@@

\begin{enumerate}
	\item Does not explicitly pick up on nucleation/zippering events. Because these are all unique sites (and zippering dynamics could vary at each site) it would be difficult to properly sample and interpret the unique corresponding modes. Furthermore these zippering dynamics occur significantly faster than slithering or GC-core fraying and cannot be captured via our save rate. 
	\item Coarse grain-models makes it difficult to compare timescales. Different degrees of freedom might have different accelerations.
	\item Limited by memory (number of frames) required to capture equilibrium trajectories and rarity of hybridization/dissociation events
	\item A smoother free energy surface minimizes the time spent in these intermediate states
	\item Sensitive to 3spn2 artefacts. Slithering may be overexpressed (and is in fact suppressed in \citep{Romano2013DNADependence}, but similar results have been shown in all-atom studies as well \citep{Xiao2019}. Base-pair mismatch interactions may be somewhat unphysical and alter state populations. Although we present slithering as a possible explanation for GC-end signal, there has been no conclusive evidence of these dynamics outside out all-atom and coarse-grained simulation studies.
\end{enumerate}

>>>>

In our analysis, we found that the AT-all sequence, given its repetitive structure and lack of GC-content, produced the cleanest dynamics and displayed a canonical spectral gap between modes. For this reason, we lead our discussion with this sequence and use it as a case study to work through our SRV-MSM pipeline step-by-step. Our first task was to identify the SRV lag time that was longer than the intrinsic Markov timescales of the system, yet short enough to resolve the dynamics of interest \citep{Phys2018MarkovValidation}. We found that most implied timescales converge at an SRV lag time of 1.2 ns. We kept a looser constraint on the convergence of the leading mode -- corresponding to the overall hybridization/dissociation process -- as it tends to have a longer transit period and becomes more difficult to fulfill Markovian conditions. It should be noted that our lag time selection was informed by the other sequences as well to maintain consistency. Next we selected an optimal number of SRV components to include in our analysis. After a certain point, higher order dynamical modes provide diminishing contributions the overall kinetic variance as measured by the Vamp-2 score, and the model can begin fitting on statistical noise in the trajectory data instead of the true dynamics \citep{McGibbon2015VariationalKinetics}. It is also more difficult to perform kmeans clustering on a high dimensional space, especially when those higher dimensions are less kinetically relevant \citep{Pande2010EverythingAsk}. For these reasons, the number of slow SRV components should be carefully selected based on the specific system of interest. As shown in (Figure \ref{fig:allseq_srv_crossval}), we see diminishing returns in the VAMP-2 score after five slow modes and select these modes as our optimized SRV basis. Next, we seek to interpret the physical relevance of these leading modes by plotting the Pearson correlation of each mode with the 100 intermolecular distances between strands. The quantitative meaning of these coordinates can be difficult to interpret given their nonlinear relationship to the SRV collective variables, but the relative different between these correlations shows which coordinates are most affected by (or effective on) each process. For example, the first slow mode shows a significant positive correlation to each distance and the strongest correlation with matching base pair distances (shown along the main diagonal). Given these relationships and the substantially longer timescale of this process, we can deduce that this leading mode corresponds to the overall hybridization and dissociation process. The next four SRV components all show a relatively high correlation along offset diagonals. These diagonals correspond to the intermolecular distances between complementary but out-of-register base pairs and point to the existence higher order "shifting" processes between sets of such base pairings. Previous "inchworm" and "pseudoknot" mechanisms have similarly been reported in simulation studies to correct base pair mismatches and occur on orders of magnitude longer timescales than underlying fast dynamics such as fraying \citep{Romano2013DNADependence, Markegard2015, Xiao2019, Maciejczyk2014DNAModel}.

@@

From this analysis we can determine that the slowest dynamics are fully characterized by the association/dissociation process and shifting behavior of out of register base pairs. Although this is qualitatively informative, we can access a more holistic picture of sequence kinetics and thermodynamics by using these SRV coordinates as a basis on which to construct an MSM. Because these coordinates are already capturing a majority of the system's kinetic variance, they serve as an ideal basis on which to group frames into microstates. We performed k-means clustering, and optimized the number of microstates at 200 by monitoring VAMP-2 score. Next, we selected an MSM lag time in a similar fashion to our SRV lag time selection process. In Figure \ref{fig:AT-all_dynamic} we compare the convergence of SRV-MSM and TICA-MSM timescales, where SRV-MSM timescales converge consistently faster and to higher values of all leading modes. This enables us to select a shorter lag time and build a higher resolution model than we could from an analogous TICA basis. Setting the MSM lag time to the same 1.2 ns we used for our SRVs, we built a Bayesian MSM to calculate transition probability matrix between each microstate. Finally, PCCA+ spectral clustering was implemented to group these microstates into macrostates that each represented a collection of metastable structures. Previous works have used a common set of microstates and/or performed manual clustering of microstates based on physical read outs from simulation data (stacking score, energies, etc) \citep{Pinamonti2017PredictingModels, PinamontiTheModels}. Although these techniques are useful for performing comparisons between sequences, we saw better results when optimizing MSMs to capture the most detail of sequence individually and thus developed an independent set of microstates and macrostates for each sequence. For AT-all, we kept to the convention of clustering into n+1 macrostates, where n is the number of slow components captured by the MSMs. To visualize these six macrostates, we project the data into the two leading TICA coordinates. Although SRVs outperform these coordinates for the purpose of MSM construction, TICA coordinates represent good high variance collective variables on which to visualize free energies and state assignments \citep{Sidky}. It is clear in Figure \ref{fig:AT-all_in_one} that the PCCA state assignments are capturing free energy minima in TICA space as independent states. After assigning these macrostates we can then calculate their relative probabilities and free energies, visualize representative molecular renderings, and estimates transition probabilities between states. In this final step, we use a minimum flux cutoff of 2e-6 in order to mitigate erroneous quick transition or skipping between states.

@@

Beyond state probabilities and free energy approximations, the macrostate MSM yields valuable discrete kinetic information in the form of transition probabilities between states. Figure \ref{fig:AT-all_in_one} shows the probability of moving from one state to another within the MSM lag time (1.2 ns). All transition probabilities are higher when moving towards a more aligned state than towards a more shifted state, suggesting that these metastable shifted states play a more significant role in facilitating the hybridization process than dissociation. Furthermore, we see equal or higher transition rates from shifted states to the dissociated state than to more aligned states, indicating that the shifting-hybridization process is frequently disrupted by complete dissociation. In particular, we observe that the transition probability from double-shifted 3' state to the dissociated state is 6x higher than to the neighboring single-shifted 3' state. Although there is an approximately equal probability of transitioning from the dissociated state to the hybridized and 5' shifted states, there is a 2x lower probability of moving to the 3' shifted states. These asymmetric kinetic effects, in addition to NN and inert tail contributions described above, may further lower the 3' shifted states populations relative to 5'. From these observations we conclude that shifted states are important in facilitating the hybridization process, but that increasingly shifted states (in particular when shifted in the 3' direction) are more likely to dissociate than to evolve into a fully aligned hybridized state. We recognize that these effects could be sensitive to the specific ion environment -- especially considering the effect of ionic strength on inert tails -- and warrant further computational and experimental investigation beyond this work.

@@

Next we examine the GC-end sequence, which add G:C caps to the same repetitive AT motifs. Based on SRV cross-validation (SI) we use the leading four slow modes to define our MSM basis.  These coordinates are qualitatively similar to those we studied for AT-all, where the three faster modes each represent some shifting dynamics. When we build the SRV-MSM for the GC-end sequence we observe a larger separation between the leading SRV-MSM timescale and the proceeding modes, which is shown in SI Figure \ref{fig:GC-end_dynamic}) by a distinct spectral gap. We observe a second spectral gap after the fourth slow modes along with a drop in the model vamp-2 score (SI Figure \ref{fig:srv_crossval}). The clustered macrostates show a similar distribution where shifted populations are notably lower. The 3' double-shifted state is no longer identified as a metastable state, and the total number of states is reduced to five (again one more than the total number of slow modes). Although these states appear similar to the AT-all shifted states, the presence of the C:T or G:A mismatches replace two intact WC bonds in the corresponding AT-all states. This substantially diminishes the stability of the GC-end shifted states, resulting in lower state populations and shorter timescales.

@@

The GC-core sequence represents a departure from the dominant shifting dynamics observed for AT-all and GC-end. Based on SRV cross-validation (SI Figure \ref{fig:srv_crossval}), we selected the first three components for our analysis. Next, we built an SRV-MSM using these first three SRV modes as a basis and proceeded along the pipeline as described above. We found that four macrostate clustering was unstable -- likely because the 3rd mode is mostly providing  information about dissociation dynamics -- so we performed PCCA+ clustering into three macrostates representing the hybridized, dissociated, and "4bp-frayed" states. Again, we found transition probabilities between states and visualized representative molecular renderings. We observe that the 4bp-frayed meta-stable state is solely composed of trajectories where both G:C core base pairs are bound and one of the adjacent A:T bonds are not. Previous work suggests that once key contacts are made, the zippering mechanism ensures that the helix will quickly form outward \citep{Romano2013DNADependence, Yin2011KineticsHybridization}. Our results indicate, however, that the relative instability of AT bonds compared the the GC-core can interrupt this process and form a longer lived frayed metastable states. This occurs during the dissociation process as well, where one half of the A:T base contacts are entirely broken for a substantial period of time before the full dissociation event occurs. We observe these events to occur with equal probability on either permutable side of the helix. The MSM transition probabilities suggest that this is key intermediate state for both the hybridization and dissociation processes, however the pathways differ between directions. We find that the hybridized state has a 5x higher probability of transitioning into the frayed state within the lag time compared to transitions from the dissociated state. This is expected considering that this state is more accessible from an already bound helix. Moreover, once oligos are in this state, they are over 10x more likely to return to the hybridized state than to fully dissociate. Thus once a dissociated to 4bp-frayed transition has occurred, it is a likely to proceed into a fully hybridized conformation. On the other hand, transitions from the hybridized to 4bp-frayed state are much more frequent and are unlikely to proceed to a fully dissociated state.

@@

For GC-core, each base pair and cutoff combination appears linear with temperature, indicating a barrierless and diffusion-driven process. As expected, the acceleration factor is also highly dependent on the choice of parameters, but, given that terminal fraying likely dominates the signal, we approximate that fraying dynamics are accelerated by about two order of magnitude relative to experiments. In contrast, GC-end is distinctly exponential, especially at base pairs closer to the core of the duplex. This behavior makes sense given the thermodynamic barrier to G:C fraying, but it also deviates the GC-end experimental linear fit. In addition to fraying, we notice that the rate of shifting increases at higher temperature. Moreover, the thermodynamic stability of these shifted conformations increases relative to that of intact duplex. For T-jumps at higher initial and final temperatures, it is possible that some portion of the starting ensemble is in a shifted state (cite Ryan's work) and that the fast dynamics are increasingly influenced by alternative "looping" G:C bonding that we observed in our diffusion map analysis above. It is less likely that shifting mechanisms would have their own distinct signals, given that these occur at far lower frequency than fraying or full-dissociation, but we believe that further experimental work is necessary to investigate this phenomenon.

@@

\begin{enumerate}
	\item Does not explicitly pick up on nucleation/zippering events. Because these are all unique sites (and zippering dynamics could vary at each site) it would be difficult to properly sample and interpret the unique corresponding modes. Furthermore these zippering dynamics occur significantly faster than shifiting or GC-core fraying and cannot be captured via our save rate. 
	\item Coarse grain-models makes it difficult to compare timescales. Different degrees of freedom might have different accelerations.
	\item Limited by memory (number of frames) required to capture equilibrium trajectories and rarity of hybridization/dissociation events
	\item A smoother free energy surface minimizes the time spent in these intermediate states
	\item Sensitive to 3spn2 artefacts. Shifting may be overexpressed (and is in fact suppressed in \citep{Romano2013DNADependence}, but similar results have been shown in all-atom studies as well \citep{Xiao2019}. Base-pair mismatch interactions may be somewhat unphysical and alter state populations. Although we present shifting as a possible explanation for GC-end signal, there has been no conclusive evidence of these dynamics outside out all-atom and coarse-grained simulation studies.
\end{enumerate}
